They're Clutching up! Team Momentum in Round-Based Esports

### 2022-08-14 paper/paper.tex
mostly done with series length analysis

Finally, I evaluate behavior when teams have two rounds left to win the
series, observing a peak in COD offensive win percentages in the 4-4
state, and no such oddity in Valorant.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, breakable]}{\end{tcolorbox}}\fi

@@

\hypertarget{description-of-call-of-duty-search-and-destroy}{%
\subsection{1 Description of Call of Duty Search and
Destroy}\label{description-of-call-of-duty-search-and-destroy}}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  One team eliminates all members of the other team prior to a bomb
  plant. (Eliminating team wins.)
\item
  The offensive team eliminates all members of the defensive team after
  a bomb plant.\footnote{Merchanics regarding the bomb:

    \begin{itemize}
    \tightlist
    \item
      The bomb can be picked up by any member of the offensive team.
    \item
      The bomb carrier is not obstructed at all by carrying the bomb
      (i.e.~movement is the same, weapon usage is the same).
    \item
      The defense does not get any visual indication for who is carrying
      the bomb.
    \item
      A bomb plant takes five seconds. The timer resets if the player
      stops planting site prior to completing it.
    \item
      A bomb defuse takes seven seconds. The timer resets if the player
      ``drops'' the bomb.
    \item
      The bomb takes 45 seconds to defuse after being planted.
    \end{itemize}} (Offense wins.)
\item
  The defensive team defuses the bomb after a bomb plant.\footnote{Often
    the defensive team will try to eliminate all team members prior to
    making the defuse, but in some cases, they may try to ``ninja''
    defuse.} (Defense wins.)
\item
  The offensive team does not make a plant by the time the round timer
  ends. (Defense wins.)
\end{enumerate}

@@

\hypertarget{data}{%
\subsection{1. Data}\label{data}}

@@

I adopt the terminology ``series'' to refer to what CoD SnD players
typically call a ``match'', so as to emulate the terminology of playoff
series in professional leagues like the National Basketball Association,
National Hockey League, and Major League Baseball. A ``game'' or a
``match'' in such leagues is analogous to a ``round'' of CoD SnD.

@@

There have been only a handful of studies of the distribution of games
played in a series, most of which assume a constant probability \(p\) of
a given team winning a game in the series, regardless of the series
state. (Mosteller 1952) observed that the American League had dominated
the National League in Major League Baseball's (MLB) World Series
matchups, implying that matchups should not modeled with \$p=0.5\$..
Mosteller proposed three approaches for identifying the optimal constant
probability value of the stronger team in the World Series, finding \$p
\textbackslash approx 0.65\$. in each case: (1) a ``method of moments'',
an approach that solves for \$p\$. from the empirical average number of
games won by the loser of the series; (2) maximum likelihood, and (3)
maximum chi-square approach.

@@

Momentum, one of most discussed topics in sports analytics, goes
hand-in-hand with a discussion of the nature of series
outcomes.\footnote{We often use use ``streaks'' and momentum
  interchangeably, but as (Steeger, Dulin, and Gonzalez 2021) note,
  momentum implies dependence between events, whereas streaking does
  not.} Two opposing fallacies are observed in the context of momentum:
the ``gambler's fallacy'' (negative recency) and ``hot hand fallacy''
(positive recency). Per {[}Ayton and Fischer (2004){]}, negative recency
is ``the belief that, for random events, runs of a particular outcome
\ldots{} will be balanced by a tendency for the opposite outcome'',
while positive recency is the expectation of observing future results
that match recent results.

@@

The general formula for the probability \(P(i)\) that a best-of-\(s\)
series lasts \(i\) rounds given constant probability \(p\) of one
team\footnote{If \(p > 0.5\), then we might say that this team is the
  better team (presumably known in hindsight)} winning each round is

\[
s_1 = \frac{s - 1}{2}, s_2 = \frac{s + 1}{2}
\]

\[
P(i) = \frac{(i - 1)!}{s_1!(i - s_2)!}(p^{s_2}(1 - p)^{i - s_2} + p^{i - s_2}(1 - p)^{s_2})
\]

resulting in Table~\ref{tbl-prob-series-lasting-i-rounds}\} for
\(s = 11\) and the naive assumption that \(p = 0.5\).

\hypertarget{tbl-prob-series-lasting-i-rounds}{}
\begin{longtable}[]{@{}rr@{}}
\caption{\label{tbl-prob-series-lasting-i-rounds}The probabilities that
a best-of-11 series lasts \(i\) rounds (\(Pr(i)\), where
\(i = {6, 7, 8, 9, 10, 11}\)) under the assumption that each team has a
50\% probability (\(p = 0.5\)) of winning each game}\tabularnewline
\toprule()
Number of rounds (\(i\)) & \(P(i)\) \\
\midrule()
\endfirsthead
\toprule()
Number of rounds (\(i\)) & \(P(i)\) \\
\midrule()
\endhead
6 & 3.1\% \\
7 & 9.4\% \\
8 & 16.4\% \\
9 & 21.9\% \\
10 & 24.6\% \\
11 & 24.6\% \\
\bottomrule()
\end{longtable}

The empirical offensive round win percentage across all rounds is
47.8\%.\footnote{Offensive round win percentage has been nearly constant
  across the three games during the CDL era: MW (2020) 47.2\%, Cold War
  (2021) 47.9\%, Vanguard (2022) 48.1\%}
Figure~\ref{fig-cod-o-round-win-prop} shows that offensive round win
rate is not quite constant, although never veers more than 10\% from
this global average.

\begin{figure}

{\centering \includegraphics{images/cod_o_round_win_prop.png}

@@

\caption{\label{fig-cod-o-round-win-prop}Offensive round win percentage
by series state}

\end{figure}

>>>>

Finally, I evaluate behavior when teams have two rounds left to win the
series, observing a peak in COD offensive win percentages in the 4-4
state, and no such oddity in Valorant.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden, enhanced, breakable, frame hidden, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

\hypertarget{description-of-call-of-duty-search-and-destroy}{%
\subsection{Description of Call of Duty Search and
Destroy}\label{description-of-call-of-duty-search-and-destroy}}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  One team eliminates all members of the other team prior to a bomb
  plant. (Eliminating team wins.)
\item
  The offensive team eliminates all members of the defensive team after
  a bomb plant.\footnote{Merchanics regarding the bomb:} (Offense wins.)
\item
  The defensive team defuses the bomb after a bomb plant.\footnote{Often
    the defensive team will try to eliminate all team members prior to
    making the defuse, but in some cases, they may try to ``ninja''
    defuse.} (Defense wins.)
\item
  The offensive team does not make a plant by the time the round timer
  ends. (Defense wins.)
\end{enumerate}

\begin{itemize}
\tightlist
\item
  The bomb can be picked up by any member of the offensive team.
\item
  The bomb carrier is not obstructed at all by carrying the bomb
  (i.e.~movement is the same, weapon usage is the same).
\item
  The defense does not get any visual indication for who is carrying the
  bomb.
\item
  A bomb plant takes five seconds. The timer resets if the player stops
  planting site prior to completing it.
\item
  A bomb defuse takes seven seconds. The timer resets if the player
  ``drops'' the bomb.
\item
  The bomb takes 45 seconds to defuse after being planted.
\end{itemize}

@@

We adopt the terminology ``series'' to refer to what CoD SnD players
typically call a ``match'', so as to emulate the terminology of playoff
series in professional leagues like the National Basketball Association,
National Hockey League, and Major League Baseball. A ``game'' or a
``match'' in such leagues is analogous to a ``round'' of CoD SnD.

\hypertarget{data}{%
\subsection{Data}\label{data}}

@@

The empirical offensive round win percentage across all rounds is
47.8\%.\footnote{Offensive round win percentage has been nearly constant
  across the three games during the CDL era: MW (2020) 47.2\%, Cold War
  (2021) 47.9\%, Vanguard (2022) 48.1\%} To contextualize our subsequent
discussion of series length and momentum, table 1 shows round win
percentages by series ``state'' (i.e.~the number of round wins by each
team prior to an upcoming round). Offensive round win rate is not quite
constant, although never veers more than 10\% from this global average.

\captionsetup[table]{labelformat=empty,skip=1pt}
\begin{longtable}{rllllll}
\toprule
 & \multicolumn{6}{c}{Offensive team's round wins} \\ 
\cmidrule(r){2-7}
Defensive team's round wins & 0 & 1 & 2 & 3 & 4 & 5 \\ 
\midrule
0 & 47.8\%
(852) & 46.6\%
(408) & 43.1\%
(216) & 43.5\%
(115) & 43.3\%
(67) & 40.5\%
(37) \\ 
1 & 48.6\%
(444) & 49.3\%
(418) & 51.5\%
(309) & 43.4\%
(205) & 43.3\%
(120) & 39.4\%
(99) \\ 
2 & 52.8\%
(218) & 48.9\%
(305) & 48.9\%
(315) & 46.6\%
(262) & 48.7\%
(189) & 42.1\%
(133) \\ 
3 & 54.5\%
(123) & 46.0\%
(200) & 49.6\%
(250) & 45.6\%
(248) & 44.4\%
(214) & 44.8\%
(174) \\ 
4 & 56.9\%
(65) & 54.5\%
(145) & 47.2\%
(193) & 44.7\%
(228) & 55.2\%
(221) & 50.5\%
(208) \\ 
5 & 47.4\%
(38) & 49.4\%
(83) & 47.1\%
(136) & 50.9\%
(175) & 45.2\%
(177) & 46.0\%
(202) \\ 
\bottomrule
\end{longtable}

@@

There have been only a handful of studies of the distribution of games
played in a series, most of which assume a constant probability \(p\) of
a given team winning a game in the series, regardless of the series
state. (Mosteller 1952) observed that the American League had dominated
the National League in Major League Baseball's (MLB) World Series
matchups, implying that matchups should not modeled with \$p=0.5\$.
Mosteller proposed three approaches for identifying the optimal constant
probability value of the stronger team in the World Series, finding
\(p \approx 0.65\). in each case: (1) a ``method of moments'' approach
in which one solves for \(p\). from the empirical average number of
games won by the loser of the series; (2) maximizing the likelihood that
the sample would have been drawn from a population in which the
probability of a team winning a game is constant across the series, and
(3) minimizing the chi-square goodness of fit estimate for \(p\).

@@

Momentum, one of most discussed topics in sports analytics, goes
hand-in-hand with a discussion of the nature of series
outcomes.\footnote{We often use use ``streaks'' and momentum
  interchangeably, but as (Steeger, Dulin, and Gonzalez 2021) note,
  momentum implies dependence between events, whereas streaking does
  not.} Two opposing fallacies are observed in the context of momentum:
the ``gambler's fallacy'' (negative recency) and ``hot hand fallacy''
(positive recency). Per (Ayton and Fischer 2004), negative recency is
``the belief that, for random events, runs of a particular outcome
\ldots{} will be balanced by a tendency for the opposite outcome'',
while positive recency is the expectation of observing future results
that match recent results.

@@

The general formula for the probability \(P_E(i)\) that a best-of-\(s\)
series lasts \(i\) rounds given constant probability \(p\) of one
team\footnote{If \(p > 0.5\), then we might say that this team is the
  better team (presumably known in hindsight)} winning each round is

\[
q = 1 - p, s_1 = \frac{s - 1}{2}, s_2 = \frac{s + 1}{2}, s_2^{'} = 1 - s_2
\]

\begin{equation}\protect\hypertarget{eq-series-length}{}{
P_E(i) = \frac{(i - 1)!}{s_1!(i - s_2)!}(p^{s_2}q^{s_2^{'}} + p^{s_2^{'}}q^{s_2}).
}\label{eq-series-length}\end{equation}

Table 1 shows the expected proportion for \(s = 11\) and the naive
assumption that \(p = 0.5\), along with the observed proportion
\(P_{O}(i)\) for CoD SnD.

\begin{table}

\caption{\label{tbl-panel}: The probabilities that a best-of-11 series
lasts \(i\) rounds (\(P(i)\), where \(i \in R = [6, 7, 8, 9, 10, 11]\))
under the assumption that each team has a 50\% probability (\(p = 0.5\))
of winning each game \{\#tbl-prob-series-lasting-i-rounds\} along with
the observed frequencies of series lasting \(i\) rounds, expressed as a
percentage \(P_O(i)\) and as a count
(\(N_O\))}\begin{minipage}[t]{\linewidth}
\subcaption{\label{tbl-panel-1}}

{\centering

\captionsetup[table]{labelformat=empty,skip=1pt}
\begin{longtable}{rrrr}
\toprule
&  & \multicolumn{2}{c}{Observed} \\ 
\cmidrule(lr){3-4}
Series lasts $i$ rounds & $P_E(i)$ & $P_O(i)$ & $N_O(i)$ \\ 
\midrule
6 & $3.1\%$ & $4.7\%$ & 40 \\ 
7 & $9.4\%$ & $11.9\%$ & 101 \\ 
8 & $16.4\%$ & $16.5\%$ & 141 \\ 
9 & $21.9\%$ & $21.7\%$ & 185 \\ 
10 & $24.6\%$ & $21.5\%$ & 183 \\ 
11 & $24.6\%$ & $23.7\%$ & 202 \\ 
\bottomrule
\end{longtable}

@@

\end{minipage}%

\end{table}

Calculating the chi-square goodness of fit statistic

\[
\chi^2 = \sum^{11}_{i=6} \frac{(P_O(i) - P_E(i))^2}{P_E(i)}, i \in R = [6, 7, 8, 9, 10, 11]
\]

as 16.0 (p-value of 0.0068), we can comfortably reject the constant
probability null hypothesis, event at a confidence level of
\(\alpha = 0.01\).

Table~\ref{tbl-mosteller-methods} shows the alternate values for the
constant round win probability \(p\) for winning a given round in a CoD
SnD that we find when applying the three methods suggested by (Mosteller
1952). Each is approximately or equal to 0.575, and each results in a
\(chi^2\) value for which we cannot reject constant probability null
hypothesis.

\hypertarget{tbl-mosteller-methods}{}
\begin{longtable}[]{@{}lrr@{}}
\caption{\label{tbl-mosteller-methods}Alternate estimates of the
constant probability \(p\) for winning a given round in a CoD SnD,
applying the three methods suggested by (Mosteller 1952), in addition to
the naive \(p = 0.5\).}\tabularnewline
\toprule()
Method & \(p\) & \(\chi^2\) (p-value) \\
\midrule()
\endfirsthead
\toprule()
Method & \(p\) & \(\chi^2\) (p-value) \\
\midrule()
\endhead
0. Naive & 0.5000 & 16.0 (\textless=0.01) \\
1. Method of moments & 0.5725 & 3.6 (0.6) \\
2. Maximum likelihood & 0.5750 & 3.5 (0.62) \\
3. Minimum \(\chi^2\) & 0.5775 & 3.5 (0.62) \\
\bottomrule()
\end{longtable}

Table~\ref{tbl-expected-series-lengths-alternative-ps} shows the new
\(P_E(i)\) when re-applying Equation~\ref{eq-series-length} for each new
\(p\). We observe that \(P_E(i)\) is notably smaller for \(p_{1,2,3}\)
when \(i \in [10, 11]\) and higher for \(i \in [6, 7]\), more closely
matching \(P_O(i)\).

\hypertarget{tbl-expected-series-lengths-alternative-ps}{}
\begin{longtable}[]{@{}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2451}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1373}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1765}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1765}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0980}}@{}}
\caption{\label{tbl-expected-series-lengths-alternative-ps}Caption}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedleft
Series lasts \(i\) rounds
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
0. \(p\)= 0.5
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
1. \(p\) = 0.5725
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
2. \(p\) = 0.575
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
3. \(p\) = 0.5775
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
\(P_O(i)\)
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedleft
Series lasts \(i\) rounds
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
0. \(p\)= 0.5
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
1. \(p\) = 0.5725
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
2. \(p\) = 0.575
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
3. \(p\) = 0.5775
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
\(P_O(i)\)
\end{minipage} \\
\midrule()
\endhead
6 & 3.1\% & 4.1\% & 4.2\% & 4.3\% & 4.7\% \\
7 & 9.4\% & 11.1\% & 11.2\% & 11.4\% & 11.9\% \\
8 & 16.4\% & 17.7\% & 17.8\% & 17.9\% & 16.5\% \\
9 & 21.9\% & 21.8\% & 21.8\% & 21.8\% & 21.7\% \\
10 & 24.6\% & 23.1\% & 23.0\% & 22.9\% & 21.5\% \\
11 & 24.6\% & 22.1\% & 22.0\% & 21.8\% & 23.7\% \\
\bottomrule()
\end{longtable}
